---
title: "Sowing Success: How Machine Learning Helps Farmers Select the Best Crops"
subtitle: "Building multi-class classification models to predict the Crop Type and identify the single most importance feature for predictive performance"
dataset_by: "Dataset provided by Datacamp"
author: "Aravindh Venkatraman"
date: "21 September 2025"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 3
    df_print: tibble
    code_folding: show
    theme: default
editor_options:
  markdown:
    wrap: 72
---

# Table of Contents {.unnumbered}

[Statements](#statements)\
[Executive Summary](#executive-summary)

1.  [Introduction](#introduction)
2.  [Aim and Methodology of this Project](#aim-and-methodology-of-this-project)
3.  [Exploratory Data Analysis - EDA](#exploratory-data-analysis-EDA)
4.  [Machine Learning Model](#machine-learning-model)
5.  [Machine Learning Model - After Feature Engineering](#machine-learning-model-after-feature-engineering)
6.  [Results & Discussion](#results-&-discussions)
7.  [Conclusion, Recommendations & Next Steps](#conclusion)


# Statements {#statements .unnumbered}

**Acknowledgement**\
I sincerely thank my parents and family for their unwavering support and encouragement, which enabled me to dedicate my time and effort to learning Machine Learning and Artificial Intelligence and applying them to environmental management challenges. I thank the Google Career Certification and Datacamp Certification courses for providing me the comprehensive resources and guidance to enhance my skills in Python, R Programming, and Machine Learning Concepts.

**Use of generative artificial intelligence**\
Generative Artificial Iintelligence (GenAI) tools were primarily used to assist in creating visualization, refining charts, and adjusting plotting parameters. In addition, GenAI was used for helping me write, debug code and improve workflow efficiency. All outputs generated by GenAI were carefully reviewed, modified and evaluated before implementation to ensure scientific accuracy, reliability, and alignment with the research objectives. 

**Ethical Consideration and Transparency**\
This research was conducted with transparency, rigor, and ethical responsibility. All computational methods and analyses were independently validated, and any use of automated or AI-assisted tools was fully disclosed. The methodologies and code are provided to ensure reproducibility and facilitate future research.

**Data and Code Availability**\
Where applicable, datasets, code scripts, and analysis workflows used in this research are openly available upon request or in public repositories, supporting transparency and reproducibility in 
computational environmental management research.

# Executive Summary {#executive-summary .unnumbered}

**Problem Statement**\
Accurately predicting .

**Project Aim and Focus**\
The project leverages machine learning techniques to evaluate predictive performance of .

**Raw data used**\
The project uses a dataset (`recipe_site_traffic_2212.csv`) which includes the following columns:

-   ``: 
-   ``: 
-   ``: 
-   ``: 
-   ``: 

Each row in this 

**Methodology**\
The project applies multi-class classification machine learning models to predict crop suitability from soil parameters. Key steps include:

-   Data Exploration and Preprocessing
-   Feature Analysis and Selection
-   Model Selection, Development and Training
-   Model Evaluation and Training
-   Feature Importance

This project demonstrates how data-driven approaches can provide
actionable insights to farmers, enabling smarter crop selection and
promoting sustainable agricultural practices.

**Results**\

**Key recommendations**\


# Introduction {#introduction}

Agriculture is the backbone of food security, yet crop planning
decisions are often based on intuition or traditional practices rather
than scientific evidence. Every growing season, farmers have the
critical decision making task of selecting the most suitable crop for
their fields without a clear understanding of how soil chemistry affects
crop performance, leading to mismatches between crop requirements and
soil properties. Such mismatches can result in reduced yields,
inefficient use of resources, and unsustainable farming practices.

Accurately predicting crop types from soil parameters — such as Nitrogen
(N), Phosphorous (P), Potassium (K), and pH — offers a more systematic
and reliable method of aligning crop choice with soil conditions. By
leveraging machine learning approaches, we can move from
experience-driven decision-making to data-driven agricultural planning,
enabling farmers to maximize yield, optimize fertilizer use, and
contribute to sustainable land management.

# Aim and Methodology of this Project {#aim-and-methodology-of-this-project}

**Aim and Focus**\
The primary objective of this project is to determine which soil
parameter most strongly influences crop type selection. To achieve this,
the project employs multi-class classification techniques, where the
target variable - crop type represents different agricultural products.

The focus areas are:

-   Identifying the single most influential soil parameter (N, P, K, pH)
    independently with the crop type.
-   Comparing model performance when using individual features versus
    combined features.
-   Extracting actionable insights that highlight the importance of
    specific soil nutrients in crop cultivation.
-   Demonstrating the role of machine learning in advancing precision
    agriculture and sustainable farming practices.

This dual focus on technical model development and practical
agricultural implications ensures that the outcomes are relevant both
academically and in real-world agricultural systems.

**Methodology**\
The project applies multi-class classification machine learning models
to predict crop suitability from soil parameters. Key steps include:

-   **Data Exploration and Preprocessing** - Understanding the feature
    distribution in the dataset, handling missing values, and preparing
    the data for modelling. - Conducted descriptive statistics and
    visualization of soil parameters. - Identified outliers in N, P, K,
    and pH, simulating real-world soil variability. - Encoded the
    categorical target variable (crop) for compatibility with machine
    learning algorithms.
-   **Feature Analysis and Selection** - Analyzing the importance and
    contribution of individual soil parameters to crop selection. -
    Analyzed the distribution and variance of each soil parameter. -
    Investigated correlations between features and crop classes. -
    Selected each feature individually to assess independent predictive
    power.
-   **Model Selection, Development and Training** - Implemented machine
    learning models suitable for multi-class classification, including:
    -   Logistic Regression (multinomial setting)
    -   Decision Trees
    -   Random Forest Classifiers
    -   Support Vector Machines (SVM)
    -   Trained separate models for each soil feature (N, P, K, pH).
    -   Built combined-feature models to evaluate improvements over
        single-feature models.
-   **Model Evaluation and Training** - Assessed performance using
    metrics appropriate for multi-class classification:
    -   Accuracy
    -   Precision, Recall, F1-score (per class and macro-averaged)
    -   Confusion Matrix for misclassification analysis
    -   Compared results across features and models to identify the most
        influential parameter.
-   **Feature Importance Extraction** - For ensemble models (Random
    Forest), calculated feature importance scores. - Cross-validated
    findings with independent feature-level predictive results.

The outcome of this study demonstrates how data-driven approaches can
empower farmers to make informed decisions about crop selection based on
soil measurements while optimizing resource use and maintaining soil
sustainability.

# Exploratory Data Analysis - EDA {#exploratory-data-analysis-EDA}

The first step of the project involves exploring the feature
distribution in the dataset `soil_measure.csv` (downloaded from the
Datacamp website and referenced on Kaggle website
[here](https://www.kaggle.com/datasets/mohamedmostafa259/soil-measures)).

**Dataset Description**

The analysis uses the dataset soil_measures.csv, which contains soil
chemical properties and corresponding recommended crops. Each row in the
dataset represents an agricultural field sample, described by the
following variables:

-   **Nitrogen (`N`):** A key nutrient promoting leafy growth and
    chlorophyll production.
-   **Phosphorous (`P`):** Vital for root development, seed formation,
    and energy transfer.
-   **Potassium (`K`):** Supports water regulation, disease resistance,
    and protein synthesis.
-   **Soil acidity/alkalinity (`pH`):** Governs nutrient availability
    and microbial activity.
-   **Crop Type (`crop`) :** A categorical field denoting the crop
    recommended for that soil condition.

This dataset provides a foundation for examining how soil chemistry
correlates with agricultural outcomes and for building models that
predict suitable crops based on soil characteristics.

This includes, - examining the range, central tendency, and variability
of each soil parameter. - identifying and handling missing or duplicate
values to ensure data quality - preparing the dataset which includes
feature scaling, encoding, and any necessary transformation for
facilitating accurate predictive modelling.

## Load required libraries

First, loading the libraries and packages in R that are needed for the
success of the `Sowing Success` project. The selected libraries provide
functions for handling data, building and performing machine learning
tasks, and visualizing results.

```{r loading libraries, message = FALSE, warning = FALSE}

# Operational libraries
library(tidyverse)    # Data manipulation and visualization

# Visualization
library(ggplot2)      # Powerful and flexible plotting 
library(kableExtra)   # Table styling
library(corrplot)     # Correlation plots

# Modelling
library(caret)        # Splitting dataset into train and test data 
library(nnet)         # Multinomial Logistic Regression Model

```

## Data Loading and Pre-processing

### Data Loading

Loading the dataset `soil_measures.csv` for exploring the dataset and
performing EDA.

```{r data loading}

path <- "D:/Study/Machine Learning/Projects/Completed projects for GitHub/Sowing-Success-How-Machine-Learning-Helps-Farmers-Select-the-Best-Crops/Data/soil_measures.csv"

# load the dataset
soil_data <- read.csv(path)

kable(head(soil_data), caption = "Soil Measurements") %>%
  kable_styling(full_width = F, position = "left")

```


















































